A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2013-10-27 jacobian_calc_CMAME.tex
Finished Sec. 1.2.1

\subsubsection{The ``complex-step'' method.}  
It is possible to approximate derivatives quite accurately with a technique based on a complex variable Taylor series expansion of a function.  This method was first described by Lyness and Moler \cite{lyness1967numerical, lyness1968differentiation} and has more recently been rediscovered \cite{squire1998using, voorhees2011complex, al2010complex} for use in engineering analysis.  The basic idea is that a model parameter can be be made to be complex and expanded in a Taylor series about a small perturbation in the imaginary plane by some arbitrary value $h$ as follows
%
\begin{equation}
f ( \beta_n + i h ) = f (\beta_n) + \frac{\partial f( \beta_n )}{\partial \beta_n} \frac{i h}{1!} + \frac{\partial f^2(\beta_n)}{\partial \beta_n^2} \frac{(i h)^2}{2!} + \frac{\partial f^3(\beta_n)}{\partial \beta_n^3} \frac{(i h)^3}{3!} + \cdots
\label{eqn:complexTaylor}
\end{equation}
%
Now we can take the imaginary part of both sides of equation (\ref{eqn:complexTaylor}) and solve for the first derivative ignoring terms of $\mathcal{O}\left ( h^2 \right)$ to yield an the estimate
%
\begin{equation}
 \frac{\partial f( \beta_n )}{\partial \beta_n} \approx \frac{\mbox{Im} \left( f (\beta_n + i h) \right)}{h}.
\label{eqn:complexFirstDeriv}
\end{equation}
%
This shows that an estimate of the first derivative of a function can be made by only utilizing one functional evaluation of a perturbed model parameter up the imaginary axis.  The step-size $h$ is arbitrary and can be made as small as practical (even to machine precision without the dangers of roundoff error as in finite-differences) to yield an accurate estimate of this derivative.  The only disadvantage of this technique is the necessity of requiring the functions to accept complex numbers as arguments. Appendix~\ref{sec:appendixA} includes a simple example that illustrates the use of the complex-step method for derivative calculation of a function.

@@

In detail, a tangent-stiffness matrix can be described as a collection of the first-order partial derivatives of a vector valued function with respect to each independent vector component of the function for a given vector.   A general mathematical formula for a tangent-stiffness matrix, $K_{ij}$,
can be expressed in indicial notation as 
%
\begin{equation} 
  K_{ij} = \left. \frac{\partial F_i}{\partial X_j}\right|_{\vec{X}_0},
\end{equation}
%
where $F_i$ is the $i^{\mbox{th}}$ component of the vector valued function, $X_j$ is the $j^{\mbox{th}}$ component of the vector argument to $F$, and a particular value of the vector is chosen to linearize about $\vec{X}_0$. One then evaluates the expression for each combination $i, j$ corresponding to a (row, column) location in the tangent-stiffness matrix. By inspection, the elements of a tangent-stiffness matrix can be estimated using any of the complex-step, AD, or common finite-difference techniques for functions $F:R^1 \rightarrow R^1$, since taking partial derivatives entails holding all but a single independent vector component of the function constant, and each element of the function can be evaluated independently of the others. \\

@@

The \emph{forward-difference} (FD) formula for calculating a tangent-stiffness matrix in the setting of a solid structural mechanics problem is
%
\begin{equation} 
  K_{ij} = \frac{\partial F_i^{int}(\vec{u} + h \hat{e}_j) - F_i^{int}(\vec{u})}{h},
\end{equation}
%
Where $K_{ij}$ is the indicial notation representation of an element of the tangent-stiffness matrix at row $i$, column $j$. The first of the two terms in the numerator is the internal force, $F_i^{int}$ evaluated in the current deformed configuration, $\vec{u}$,  plus a small perturbation $h$ in the direction $\hat{e}_j$ where $\hat{e}_j$ represents a unit-vector corresponding to the $j^{\mbox{th}}$ degree-of-freedom. The second of the two terms is $F_i^{int}$ evaluated in the current configuration. The denominator is the magnitude of perturbation called \emph{probe-distance}.

The \emph{central-difference} (CD) formula for calculating a tangent-stiffness matrix is
%
\begin{equation} 
  K_{ij} = \frac{\partial F_i^{int}(\vec{u} + h \hat{e}_j) - F_i^{int}(\vec{u} - h \hat{e}_j)}{2 h}.
\end{equation}
%
Together, the FD and CD methods, are sometimes termed \emph{finite-difference probing} techniques for tangent-stiffness calculation in literature \cite{ref-Adaggio}.

The complex-step (CS) formula for calculating a tangent-stiffness matrix is
%
\begin{equation} 
  K_{ij} = \frac{\mbox{Im}(F_i^{int}(\vec{u} + i h \hat{e}_j))}{h}
\end{equation}

Where our model $F$ is treated as function of complex variables and $x$ is treated as a vector of
complex values. The expression says that the real valued partial derivative of the $i$'th component of $F$
w.r.t the $j$'th component of $x$ is equal to the imaginary component of $F_i$ evaluated at $x$
where $x$ is perturbed by a very small amount in the direciton of the imaginary axis corresponding
to the $j$'th component of $x$ all divided by the magnitude of the perturbation. There is no partial
notation because a partial difference is not taken.

@@

\begin{equation} K_{ij} = \frac{\partial F_i^{int}}{\partial x_j} \end{equation}

>>>>

\subsubsection{The ``complex-step'' method.}  
\label{sec:CSmethod}
It is possible to approximate derivatives quite accurately with a technique based on a complex variable Taylor series expansion of a function.  This method was first described by Lyness and Moler \cite{lyness1967numerical, lyness1968differentiation} and has more recently been rediscovered \cite{squire1998using, voorhees2011complex, al2010complex} for use in engineering analysis.  The basic idea is that a model parameter can be be made to be complex and expanded in a Taylor series about a small perturbation in the imaginary plane by some arbitrary value $h$ as follows
%
\begin{equation}
f ( \beta_n + i h ) = f (\beta_n) + \frac{\partial f( \beta_n )}{\partial \beta_n} \frac{i h}{1!} + \frac{\partial f^2(\beta_n)}{\partial \beta_n^2} \frac{(i h)^2}{2!} + \frac{\partial f^3(\beta_n)}{\partial \beta_n^3} \frac{(i h)^3}{3!} + \cdots
\label{eqn:complexTaylor}
\end{equation}
%
Now we can take the imaginary part of both sides of equation (\ref{eqn:complexTaylor}) and solve for the first derivative ignoring terms of $\mathcal{O}\left ( h^2 \right)$ to yield an the estimate
%
\begin{equation}
 \frac{\partial f( \beta_n )}{\partial \beta_n} \approx \frac{\mbox{Im} \left( f (\beta_n + i h) \right)}{h}.
\label{eqn:complexFirstDeriv}
\end{equation}
%
This shows that an estimate of the first derivative of a function can be made by only utilizing one functional evaluation of a perturbed model parameter up the imaginary axis.  The step-size $h$ is arbitrary and can be made as small as practical (even to machine precision without the dangers of roundoff error as in finite-differences) to yield an accurate estimate of this derivative.  The only disadvantage of this technique is the necessity of requiring the functions to accept complex numbers as arguments. Appendix~\ref{sec:appendixA} includes a simple example that illustrates the use of the complex-step method for derivative calculation of a function.

@@

In detail, a tangent-stiffness matrix can be described as a collection of the first-order partial derivatives of a vector valued function with respect to each independent vector component of the function for a given vector.   A general mathematical formula for a tangent-stiffness matrix, $K_{ij}$,
can be expressed in indicial notation as 
%
\begin{equation} 
  K_{ij} = \left. \frac{\partial F_i}{\partial X_j}\right|_{\vec{X}_0},
\end{equation}
%
where $F_i$ is the $i^{\mbox{th}}$ component of the vector valued function, $X_j$ is the $j^{\mbox{th}}$ component of the vector argument to $F$, and a particular value of the vector is chosen to linearize about $\vec{X}_0$. One then evaluates the expression for each combination $(i, j)$ corresponding to a (row, column) location in the tangent-stiffness matrix. By inspection, the elements of a tangent-stiffness matrix can be estimated using any of the complex-step, AD, or common finite-difference techniques for functions $F:R^1 \rightarrow R^1$, since taking partial derivatives entails holding all but a single independent vector component of the function constant, and each element of the function can be evaluated independently of the others. \\

@@

The \emph{forward-difference} (FD) formula for calculating a tangent-stiffness matrix in the setting of a solid structural mechanics problem is
%
\begin{equation} 
  K_{ij} = \frac{F_i^{int}(\vec{u} + h \hat{e}_j) - F_i^{int}(\vec{u})}{h},
\end{equation}
%
Where $K_{ij}$ is the indicial notation representation of an element of the tangent-stiffness matrix at row $i$, column $j$. The first of the two terms in the numerator is the internal force, $F_i^{int}$ evaluated in the current deformed configuration, $\vec{u}$,  plus a small perturbation $h$ in the direction $\hat{e}_j$ where $\hat{e}_j$ represents a unit-vector corresponding to the $j^{\mbox{th}}$ degree-of-freedom. The second of the two terms is $F_i^{int}$ evaluated in the current configuration. The denominator is the magnitude of perturbation called \emph{probe-distance}.

The \emph{central-difference} (CD) formula for calculating a tangent-stiffness matrix is
%
\begin{equation} 
  K_{ij} = \frac{F_i^{int}(\vec{u} + h \hat{e}_j) - F_i^{int}(\vec{u} - h \hat{e}_j)}{2 h}.
\end{equation}
%
Together, the FD and CD methods, are sometimes termed \emph{finite-difference probing} techniques for tangent-stiffness calculation in literature \cite{ref-Adaggio}.

The complex-step (CS) formula for calculating a tangent-stiffness matrix is
%
\begin{equation} 
  K_{ij} = \frac{\mbox{Im}(F_i^{int}(\vec{u} + i h \hat{e}_j))}{h},
\end{equation}
%
where the internal force, $F^{int}$ is now treated as function of complex vectors,  $\vec{u}$. The small perturbation, $h$, is now carried out on the imaginary axis and only the real coefficient to the imaginary part of the vector returned by $F^{int}$ is kept, hence the $\mbox{Im}(\cdot)$ operation. As highlighted in Section~\ref{sec:CSmethod}, notice there is no subtraction operation taking place; therefore $h$ can be made very small to yield highly accurate derivatives.

@@

\begin{equation} K_{ij} = \frac{\partial F_i^{int}(\vec{u})}{\partial u_j} \end{equation}
